Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2010-11-07 pkg/paper/Paper/draft.Rnw
updated Draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@71 b679b2f7-6539-45d0-a605-5ddcb976cde9

<<preliminaries>>=
## packages
Sys.setlocale("LC_TIME", "English")
library("glogis")
library("fxregime")
library("lattice")
library("ggplot2")

@@

## sequentially load data
for(i in seq_along(fnam)) {
  load(file.path("..", "Analysis", fnam[i]))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}

@@

@@

\Abstract{
  The aim of this paper is to investigate how a countries decision to join one or all of the steps leading towards the common European currency changed its inflation rate dynamics. Our attempt is to model the inflation rates via a Generalized Logistic Distribution and then test for breaks in the parameters of mean, variance and skewness.
}

@@

Theory unfortunately is quite unsure  whether or not the creation of a monetary union between two or more states is likely to reduce or increase the variability or even the level of the inflation rate.  \citet{cooper} points out, that ``a central bank under a monetary union will internalize the interdependence between countries and optimally choose a lower inflation rate" and he argues that a ``central Bank governing the growth of money supply will optimally choose zero inflation." This is not the case with the ECB which targets at 2\% so as to avoid the alleged risk of deflation. It is thus not quite clear how a monetary union will affect the volatility and the level of the inflation rate.\\

@@

The effects of joining a monetary union on inflation rate variability depend on structural parameters like risk aversion, price flexibility, export demand elasticity, openness and shock correlations. Due to the fact that not all of these parameters are known and that their interaction as well has to be estimated, theory has some troubles establishing a sound theoretical model. \\

@@

In a paper examining structural convergence of the inflation rates in EU countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that convergence in time of inflation dynamics was only partly observable. 

@@

Inquiring into the convergence properties of inflation rates among countries of the EMU, \citet{busetti} find that from 1980---1997 there was convergence of inflation rates, but afterwards there is some diverging behavior. \\

@@

Inflation is measured as the logarithm of the monthly change in the HICP from 1.1990---3.2010. Countries included are Austria,	Belgium, the	Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	the Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	the	United Kingdom.\footnote{Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity.} The data is obtained from the OECD Statistics.\footnote{http://www.oecd-ilibrary.org/content/data/data-00285-en} \\
The countries can be divided into three different groups: the EURO countries (Austria, Belgium, Estonia\footnote{although Estonia has not yet introduced the EURO, but will do it in 2011}, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, the Netherlands, Portugal, Slovakia, Slovenia and Spain)\footnote{Cypria, Malta, Andorra, Monaco, San Marino and the Vatican are left out due their minor importance.}, EU members not participating in the ERM~II (Exchange Rate Mechanism): the Czech Republic, Hungary, Poland, the United Kingdom and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU.
% alternative link: http://stats.oecd.org/index.aspx

@@

Our main question of interest is whether or not a countries decision to join one or all of the steps leading towards a common European currency changed its inflation rate dynamics. The economic reason why this should be so is rather clear cut for more than half of the countries in our sample. We would expect changes in the dynamics of all Eastern European countries due to their transition from previously centrally planned economies towards free market economies. All of these countries experienced some problems with very high inflation rates at the beginning and they all were successfull in eliminating excess inflation afterwards, it can be safely assumed that the former member countries of Yugoslavia follow the same pattern. In the case of the Southern European countries, with a history of high inflation and less pronounced fiscal discipline, we would also expect inflation dynamics to change towards a regime of lower inflation. The case for the Central European countries of course is different. There is no reason why the European Central Bank (ECB) should be more bent on fighting inflation than the former Bundesbank. Also there is no real theoretical evidence, whether or not a monetary union of former high and low inflation countries would lead to a lower inflation rate in countries with allready low inflation rates. This lack of theory of course poses a great challenge. \\
% [allthough it should be added, that the 90ies are perhaps a bad starting point, since we would expect them to be allready rather moderate compared with past values of 60ies and 70ies]
The approach we use follows from one simple idea, i.e. we want to track the evolution of mean, variance and skewness of our series over time. Thereby we hope to be able to find and estimate possible breakpoints in the data. The correlation and the correlation structure of the time series themselves are of no interest and are treated as a nuisance term.\\
To do this, we assume that our data is a draw from a Generalized Logistic Distribution (GL) and we proceed by estimating the parameters of the GL--distribution via maximum likelihood. We then test for structural change and estimate them using dynamic programming.

@@

In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tails, which fits these problems somewhat better. The use of the generalized logistic distribution as we will use it in this paper is rather rare. \citet{won} use a GL distribution in a regression model with autocorrelated errors. They use this type of distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  nonsymmetric and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyse extreme risk and value--at--risk in the German stock market, all though they don't use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in relationship with expected inflation. \citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\
We apply the general framework, as developed in \citet{z07}, to a more specific model, in this case by means of a GL distribution. Prior to that, we would like to give a short justification for the usage of the GL distribution. Regarding the data at hand, it was not possible to use the already existing method developed in \citet{z07}, since almost all inflation rates, with the notable exception of Greece, are not normally distributed and clearly exhibit asymmetric properties. \\
Therefore, a somewhat more flexible distribution had to be used. We needed a distribution exhibiting rather fatter tails and the property to be both left and right skewed. To serve this purpose, we use a generalization of the logistic distribution as defined in \citet{johnson}. For return series $y_t = 100\cdot log(HICP_t/HICP_{t-1})$ ($t = 1, \dots, n$) we assume a GL distribution given by:

@@

with location ($\theta$), scale ($\sigma$) and shape ($\delta$). For $\delta$=1 the distribution simplifies to the logistic distribution, for $\delta$$<$1 it is skewed to the left and for $\delta$$>$1 it is skewed to the right. The moments are given by:

@@

where $\gamma()$ is the digamma function and its derivatives. \\
The log-likelihood is given by:

@@

The resulting score function has 3 components $(s_{\theta}, s_{\sigma}, s_{\delta})$, with $\tilde{y} = \exp^{-\frac{y-\theta}{\sigma}}$, where the scores are the derivatives of the log--likelihood function and sum up to 0. These are used later on for the construction of the test statistics and are a means of measuring the estimation error in a maximum likelihood framework.

@@

An enhancement to the already existing \proglang{R} package \pkg{strucchange}, which currently does not support a generalized logistic distribution (GL) is provided with this paper. The asymptotic testing theory still holds for this generalization. \\
The first part of the results we are going to present here, i.e., the tests and the graphical illustration of the empirical fluctuation process goes back to \citet{z07}. the second part of the results - the dating procedure and the illustration of the densities fitted for the subsamples (divided by the breaks) - ca be found in \citet{z10}.

@@

We then proceed step by step. First, we estimate the prameters of the GL--distribution by means of maximum likelihood, which is possible, if we assume that $\phi = (\theta, \sigma, \delta)$ is stable over time $t$:

@@

Our problem now simplifies to the following question: Is the assumption valid or do the $\phi_t$ vary over time?

@@

This can be assessed using the empirical scores $s(y_t|\hat{\phi})$, as given above, as measures of model deviation.  The empirical scores process $\mathit{efp}(\cdot)$ captures deviations from zero over time and is given by:

@@

where $\hat{V}^{-1/2}$ is some consistent estimator for the variance of the scores, it is used to decorrelate the series as to avoid problems due to serial correlation. We then employ a Functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ which converges to a 3-dimensional Brownian bridge:

@@

Lets elaborate on this basic idea. Consider some time series given by the following graph. We would not need any test, since we see a clear break at the middle of the sample.

@@

If we  estimate a simple least squares regression we would get a straight regression line at the mean of the series.  What we do now is to sum up the scores (which can be viewed as similar to the residuals) by means of the efp (i.e. something like a cumulated sum of residuals). This summing up of the resiudals would lead to first summing up all the negative residuals down to the lowest point, then  we proceed by summing up the other (positive) residuals. The peak of this cumulated sum will be our guess of the breakpoint. If this peak crosses some test statistics, then it can be viewed as a significant deviation of our hpyothetized stable parameters.

@@

The supLM test is the single most powerful test for a single abrupt change in unknown time and has the highest power if we suspect a change in more than one parameter of our distribution. Throughout the paper we will use just the supLM test.

@@

All parameters $\tau_1, \dots, \tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic programming.

@@

To help with the interpretation, we give a short overview over the history of the EMU. Following the Delors Plan with his 3 stages, we have: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2002) the transition period, which ended with the introduction of the Euro as legal tender and started with the Exchange Rate Mechanism~II (ERM~II) for Non--members and the irrevocable fixing of exchange rates of Member States. \\
It should be pointed out however, that these monthly series do still contain all the saisonality and that if we correct for this by using the lag of order 12 instead of order 1, we find somewhat less pronounced changes with the exception of the latest observations, with the financial crises and the overall lowering in inflation rates, which of course leads to a considerable increase in variance for more than half of our sample. 5 countries in our sample then show a break in October 2007, coinciding with the ongoing financial crises.

@@

%If the EMU had a significant effect upon inflation rate dynamics, we would expect to find a break either in the 90ies, reflecting the various waves of integration or after the introduction of the Euro due to the popular argument that the introduction of Euro paper money led to a considerable price increase. \\
%[some considerations: 1. necessary to include all the results? seems to me preferable to deal with them as groups of individual contries rather than all on one page; 2. show all the groups? 3. show example like in Slovenian case?]

@@

On the basis of these results, we are able to trace out different groups. The first of this group consits of Belgium and Luxembourg. This may come as no surprise since both countries formed a monetary union prior even to the start of the Exchange Rate Mechanism (ERM). While both countries enjoyed very stable inflation rates with low values of both mean and variance, this changed around the timing of the irrevocable fixing of the exchange rates. A causal connection to this events seems possible. However, according to an official of Statec Luxembourg\footnote{www.statec.lu}, these changes can be traced back  to the inclusion of sales (winter and summer) from january 1999 on according to a change in regulation. For Belgium we suspect similar change, but our suspicion is not yet confirmed.

% Quelle: e-mail Verkehr mit Jérôme Hury, dem Chef der Statec Price Abteilung

% interesting thing: both were allready in a currency union (the belgish--luxembourg union)
% Luxembourg and probably in Belgium too, as  you find out sales prices were excluded before the regulation

@@

The second of these groups are the Eastern European Countries, the Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90ies. Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later.

@@

Another combination of countries with -- this time -- identical results are the two biggest southern European countries Italy and Spain. Both of them experienced high rates of inflation in the early 90ies. Afterwards, in the phase leading up to the monetary union, mean and variance were significantly reduced. This trend changed drastically soon after the fixing of the exchange rates leading to higher mean and much higher variances then ever before since the beginning of the HICP series.

@@

Ireland can be considered a rather special case. It is the only economy where the effects of the ongoing financial crisis of 2008 are visible. We find a structural break in 2008 following the strong contraction of the Irish housing bubble and a beginning deflation.

@@

Another interesting group consists of 3 economically very different countries. In all of them we find no evidence of a structural change. Whereas the Netherlands and Finland are both on the lower side of the inflation margin, Greece exhibits some very strange behaviour. It has by far the highest variance of all series and seems to be completely different from all the others, just like some randomly generated series.

@@

The last group with clear patterns can be seen with the two countries that decided not to be part of any European Monetary System\footnote{If we ignore the short participating time of the United Kingdom in the EMS}, the United Kingdom and Sweden. Both have a clear break in the early 90ies, and both of these breaks can be traced back to crisis, the currency crisis of the United Kingdom with the "`Black Wednesday"' in September 1992 and the run on the Swedish currency in 1992.

@@

No clear message is available for: Austria, Denmark, France, Germany and Portugal.

@@

\subsection{An example--Slovenia}

Let us go through all the steps of our analysis by means of an example. An ideal example for this is the case of Slovenia. The economic reason why we suppose to find changing dynamics are clear. In the first phase of the consolidation after the break--up of Yugoslavia and after the short war of 1991, Slovenia experienced very high inflation rates, 
with yearly growth rates as high as 90\% in August 1990. But gradually Slovenia was able to reduce the inflation rate towards much lower values. This first phase is not captured in our data since it starts as late as January 1996 for Slovenia. For the period up to today we would expect one or two breaks from an economic point of view. The efforts to reach the Maastricht criteria as soon as possible are likely to lead to a further lowering of the mean and possibly also of the variance of the inflation rate. Another break point might possibly be found towards the introduction of the EURO coins themselves in 2007. \\
Actually when we look at the data, this is precisely what we get:

<<>>=
i <- 18
nam <- colnames(HICP)[i]
fnam <- paste(nam, ".rda", sep = "")
@
%\section*{\Sexpr{nam}}
<<>>=
if(file.exists(fnam)) {
load(fnam)
} else {
<<analysis>>
}
@
\begin{center}
<<slovplot, fig=TRUE, height=4, width=8>>=
<<plot1>>
@
\end{center}

To be as precise as possible with break points, we also include a graph showing the values of the LWZ (Liu, Wu and Zidek) criteria as developed in \citet{liu}. Whereas the supLM test suggest at least one break, the LWZ criterion favors no break. But the decline in the negative log--likelihood from 0 breaks to 1 is the strongest, we concluded that one break point actually would be the best alternative.

@@

We find a break in August 2003 d with a much lower mean value (only a third now of its previous value) and interestingly enough, a higher volatility. The supLM test graph on the right hand side suggests one break around 2003 and possibly another towards the end of the sample. This second break point is not estimated, since we restricted a regime towars a length of at least 24 month. \\
Slovenia did achieve this by two very important measures. First it liberalized the financial markets in 2003 and -- most important -- it restricted the year--on--year growth of the monetary base M1 severely starting in 2003. Combined these were enough to ensure  Slovenian participation in the EURO--zone.

@@

>>>>

<<preliminaries>>=
## packages
if(Sys.getlocale() != "C") Sys.setlocale("LC_TIME", "English")
library("glogis")
library("fxregime")
library("lattice")

@@

## sequentially load data
## old: load(file.path("..", "Analysis", fnam[i]))

for(i in seq_along(fnam)) {
  load(file.path(root_dir, "Analysis", fnam[i]))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}

@@

xyplot.gefp <- function(x, panel = NULL, ylim = NULL, ...)
{
  cval <- maxBB$computeCritval(0.05, ncol(x$process))
  if(is.null(ylim)) ylim <- range(c(range(x$process), 1.1 * cval, -1.1 * cval))

  gefp_plot <- function(x, y, subscripts, groups, panel = panel.xyplot,
    col = 1, type = "p", pch = 20, lty = 1, lwd = 1, ...)
  {
    col <- rep(as.list(col), length = nlevels(groups))
    type <- rep(as.list(type), length = nlevels(groups))
    pch <- rep(as.list(pch), length = nlevels(groups))
    lty <- rep(as.list(lty), length = nlevels(groups))
    lwd <- rep(as.list(lwd), length = nlevels(groups))

    for(g in 1:nlevels(groups)) {
      idx <- levels(groups)[g] == groups[subscripts]
      if (any(idx)) panel(x[idx], y[idx], ...,
        col = col[[g]], type = type[[g]], pch = pch[[g]],
        lty = lty[[g]], lwd = lwd[[g]])
      grid::grid.lines(y = grid::unit(0, "native"))
      grid::grid.lines(y = grid::unit(cval, "native"), gp = grid::gpar(col = 2))
      grid::grid.lines(y = grid::unit(-cval, "native"), gp = grid::gpar(col = 2))
    }
  }
  if(is.null(panel)) panel <- gefp_plot

  xyplot(x$process, panel = panel, ylim = ylim, ...)
}

@@

\Abstract{
  The aim of this paper is to investigate how a countries decision to join one or all of the steps leading towards the common European currency changed its inflation rate dynamics. Our attempt is to model the inflation rates via a Generalized Logistic Distribution (GL) and then test for breaks in the parameters of the distribution.
}

@@

Theory unfortunately does not provide any clear foundation for a proper answer whether or not the creation of a monetary union between two or more states is likely to reduce or increase the variability or even the level of the inflation rate.  \citet{cooper} points out, that ``a central bank under a monetary union will internalize the interdependence between countries and optimally choose a lower inflation rate" and he argues that a ``central Bank governing the growth of money supply will optimally choose zero inflation." This is not the case with the ECB which targets at 2\% so as to avoid the alleged risk of deflation. It is thus not quite clear how a monetary union will affect the volatility and the level of the inflation rate.\\

@@

The effects of joining a monetary union on inflation rate variability depends on structural parameters like risk aversion, price flexibility, export demand elasticity, openness and shock correlations. Due to the fact that not all of these parameters are known and that their interaction as well has to be estimated, the whole model is very dependend on assumptions. \\

@@

In a paper examining structural convergence of the inflation rates in EU countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that convergence over time of inflation dynamics was only partly observable. 

@@

Inquiring into the convergence properties of inflation rates among countries of the EMU, \citet{busetti} find that from 1980---1997 there was convergence of inflation rates, but afterwards there is some diverging behavior. This does not quite go along with our findings, that do not indicate stronger divergence after 1997 by means of the detection of break points. \\
In section 2 we will present the data, in section 3 we will present our approach, in section 4 we will present the results, in section 5 we will discuss our findings and try to bring them into context with the overall economic condition, section 6 concludes.

@@

Inflation is measured as the monthly change in the HICP rates from 1.1990---3.2010. Countries included are Austria,	Belgium, the	Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	the Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	the	United Kingdom.\footnote{Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity.} The data is obtained from the OECD Statistics.\footnote{http://www.oecd-ilibrary.org/content/data/data-00285-en} \\
The countries can be divided into three different groups: the EURO countries (Austria, Belgium, Estonia\footnote{although Estonia has not yet introduced the EURO, but in 2011.}, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, the Netherlands, Portugal, Slovakia, Slovenia and Spain)\footnote{Cypria, Malta, Andorra, Monaco, San Marino and the Vatican are left out due their minor importance.}, EU members not participating in the ERM~II (Exchange Rate Mechanism): the Czech Republic, Hungary, Poland, the United Kingdom and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU.
% alternative link: http://stats.oecd.org/index.aspx

@@

Our main question of interest is whether or not a countries decision to join one or all of the steps leading towards a common European currency changed its inflation rate dynamics. The economic reason why this should be so is rather clear cut for more than half of the countries in our sample. We would expect changes in the dynamics of all Eastern European countries due to their transition from previously centrally planned economies towards free market economies. All of these countries experienced some problems with very high inflation rates at the beginning and they all were successfull in eliminating excess inflation afterwards. It can be safely assumed that the former constituent republics of Yugoslavia follow the same pattern. In the case of the Southern European countries, with a history of high inflation and less pronounced fiscal discipline, we would also expect inflation dynamics to change towards a regime of lower inflation. The case for the Central European countries of course is different. There is no reason why the European Central Bank (ECB) should actually  follow the rather stable monetary policies of the former German central bank in a one--to--one fashion. \\
% [although it should be added, that the 90ies are perhaps a bad starting point, since we would expect them to be already rather moderate compared with past values of 60ies and 70ies]
The approach we use follows one simple idea: i.e. we want to track the evolution of mean, variance and skewness of our series over time. Thereby we hope to be able to find and estimate possible breakpoints in the data. The correlation and the correlation structure of the time series themselves are of no interest and are treated as a nuisance term.\\
To do this, we assume that our data is a draw from a Generalized Logistic Distribution (GL) and we proceed by estimating the parameters of the GL--distribution via maximum likelihood. We then test for structural change and estimate them using dynamic programming.

@@

In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tails, which fit these kind of data somewhat better. The application of the generalized logistic distribution as we will use it in this paper is rather rare. \citet{won} use a GL distribution in a regression model with autocorrelated errors. They use this type of distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  non symmetric and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyze extreme risk and value--at--risk in the German stock market, although they don't use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in context to expected inflation. \citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\
We apply the general framework, as developed in \citet{z07}, to a more specific model, in this case by means of a GL distribution. Prior to that, we would like to give a short justification for the usage of the GL distribution. Regarding the data at hand, it was not possible to utilize the already existing method developed in \citet{z07}, since almost all inflation rates, with the notable exception of Greece, are not normally distributed and clearly exhibit asymmetric properties. \\
Therefore, a somewhat more flexible distribution had to be used. We needed a distribution exhibiting rather fatter tails and the property to be both left and right skewed. To serve this purpose, we use a generalization of the logistic distribution as defined in \citet{johnson}.

\clearpage
For our inflation series $y_t = 100\cdot log(HICP_t/HICP_{t-1})$ we assume a GL distribution given by:

@@

with location ($\theta$), scale ($\sigma$) and shape ($\delta$). For $\delta$=1 the distribution simplifies to the logistic distribution, for $\delta$$<$1 it is skewed to the left and for $\delta$$>$1 it is skewed to the right. The moments are:

@@

where $\gamma(\cdot)$ is the digamma function and its derivatives. \\
The log-likelihood is:

@@

The resulting score function, where the scores are the derivatives of the log--likelihood function, has 3 components $(s_{\theta}, s_{\sigma}, s_{\delta})$, with $\tilde{y} = \exp^{-\frac{y-\theta}{\sigma}}$. These are used later on for the construction of the test statistics and are a means of measuring the estimation error in a maximum likelihood framework.

@@

An enhancement to the already existing \proglang{R} package \pkg{strucchange}, which currently does not support a GL distribution is provided with this paper. The asymptotic testing theory still holds for this generalization. \\
The first part of the results we are going to present here, i.e., the tests and the graphical illustration of the empirical fluctuation process can be backtracked to \citet{z07}, the second part of the results - the dating procedure and the illustration of the densities fitted for the subsamples (divided by the breaks) - can be found in \citet{z10}.

@@

In order to analyze our data, we first estimate the parameters of the GL--distribution by means of maximum likelihood, which is possible, if we assume that $\phi = (\theta, \sigma, \delta)$ is stable over time $t$:

@@

Our problem now simplifies to the following question: Is the assumption valid or do the $\phi_t$ change?

@@

This can be assessed using the empirical scores $s(y_t|\hat{\phi})$, as presented above, as measures of model deviation.  The empirical scores process $\mathit{efp}(\cdot)$ captures deviations from zero over time and is given by:

@@

where $\hat{V}^{-1/2}$ is some consistent estimator for the variance of the scores, which is used to decorrelate the series as to avoid problems due to serial correlation. We then employ a Functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ which converges to a 3-dimensional Brownian bridge:

@@

\clearpage
Lets elaborate on this basic idea. Consider some time series given by the following graph. We would not need any test, since we see a clear break at the middle of the sample.

@@

If we  estimate a simple least squares regression we get a straight regression line at the mean of the series.  What we do now is to sum up the scores (which can be viewed as similar to the residuals) by means of the efp (i.e. something like a cumulated sum of residuals). This summing up of the residuals would be constructed by first aggregating all the negative residuals down to the lowest point, then  we proceed by accruing the other (positive) residuals. The bottom of this cumulated sum will be our estimation of the breakpoint. If this peak crosses some test statistics, then it can be viewed as a significant deviation from the null hypothesis.

@@

The supLM test is the most powerful test for a single abrupt change in unknown time and has the highest power if more than one parameter of the distribution changes over time. Throughout the paper we will use the supLM test.

@@

All parameters $\tau_1, \dots, \tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic programming. More details for this estimation technique are provided in \cite{z10}.

@@

To help with the interpretation, we give a short overview over the history of the EMU. It basically consisted of 3 stages: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2002) the transition period, which ended with the introduction of the Euro as legal tender and started with the Exchange Rate Mechanism~II (ERM~II) for Non--members and the irrevocable fixing of the exchange rates of the Member States. \\
It should be pointed out however, that these monthly series do still contain all the seasonality\footnote{We would like to mention that in the case of Belgium and Luxembourg, the seasonality patterns can be traced back to changes in HICP definition and legal proquirements. In addition to Belgium and Luxembourg, there are a number of other states, where the sales periods are regulated by law. France has an individual time span regulated by each department. This should pose no problem, since there is no global state level solution. In Greece there are also two set sales--periods in winter and summer. Given the probelems concerning the Greek data, this does not pose any addional difficulty. In Italy the sales periods are fixed each year by the Chambers of Commerce. Portugal has two set sales periods as well, but for two month (January and February as well das Mid--July to Mid--September. The source for this information is: $http://www.konsumenteuropa.se/en/Shopping-in-Europe$}. If we correct for this by using the lag of order 12 instead of order 1, we find somewhat less pronounced changes with the exception of the latest observations, with the financial crises and the overall lowering in inflation rates, which of course leads to a considerable increase in variance for more than half of our sample. 5 countries change their inflation dynamics after October 2007, coinciding with the ongoing financial crises.

@@

%If the EMU had a significant effect upon inflation rate dynamics, we would expect to find a break either in the 90ies, reflecting the various waves of integration or after the introduction of the Euro due to the popular argument that the introduction of Euro paper money led to a considerable price increase. \\
%[some considerations: 1. necessary to include all the results? seems to me preferable to deal with them as groups of individual countries rather than all on one page; 2. show all the groups? 3. show example like in Slovenian case?]

@@

On the basis of these results, we are able to trace out different groups. The first of this groups consists of Belgium and Luxembourg. This may come as no surprise since both countries formed a monetary union prior to the start of the Exchange Rate Mechanism (ERM). While both countries enjoyed very stable inflation rates with low values of both mean and variance, this changed around the timing of the irrevocable fixing of the exchange rates. But in the case of Belgium and Luxembourg, what we actually observe is a change due to the definition of the HICP. According to an official of the OECD, these changes can be traced back  to the inclusion of sales (winter and summer) in January 1999 for Belgium and January 2000 for Luxembourg which was demanded by a Commission regulation.\footnote{for technical details see: http://eur-lex.europa.eu/LexUriServ/LexUriServ.do?uri=OJ:L:2000:300:0016:0017:EN:PDF} This lead to a 6 month seasonality pattern (January and July) afterwards (meaning after the breakpoints induced by sales inclusion), since sales periods in Belgium and Luxembourg are regulated by law and have to occur in the first two weeks of these months and therefor we observe this sharp price decreases in both month. Interestingly enough, both these price decreases are immediately followed by increases in February and August of even higher magnitude.

%%%%% Seasonality Problems:
% for the others: http://www.konsumenteuropa.se/en/Shopping-in-Europe/Luxembourg/
% we do not find many more regulated sales periods in Europe (given the data above is correct, which I hope :)
% exemption: France (regulated by each department individually (so changes hopefully leveled out) --> no problem
% 	         Greece (In Greece there are two set sales-periods. Winter sales: usually during January – February. Summer sales: usually during July – August.)
%            --> no problem, dont trust series anyway
%            Italy (There are two set sales periods in Italy, in winter and summer. Dates and duration are fixed by the Chambers of Commerce.)
%            --> what if these changes???  --> e-mail sent to italian chamber of commerce in austria, awaiting reply
%            Portugal (There are two set sale periods: Winter sales: 28/12 to 28/02 Summer sales: 15/07 to 15/09 )
% Quelle für diese Änderung : e-mail Verkehr mit Jérôme Hury, dem Chef der Statec Price Abteilung; und Anne-Sophie.FRAISSE@oecd.org

@@

The second of these groups are the Eastern European countries, the Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90ies. Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later.

@@

Another combination of countries with -- this time -- identical results are the two biggest southern European countries Italy and Spain. Both of them experienced high rates of inflation in the early 90ies. Afterwards, in the phase leading up to the monetary union, mean and variance were significantly reduced. This trend changed drastically soon after the fixing of the exchange rates leading to higher mean and much higher variances then ever before since the beginning of the HICP series. \\
Italy is quite a nice example to explain the usefullness of the skewness parameter. If we focus on the first two periods, with skewness lower than 1 (which indicates that the distribution is skewed to the right) we find that positive shocks are not normally neutralized by negative shocks. This changes in the last phase with skewness negative, indicating that the GL is now skewed to the left and that positive shocks will more likely be followed by negative shocks.

% skewness: if delta = 1.1 --> skweness is 0.1, is positive, skewed to the right: linkssteil
% ie. negative skew (delta <1: rechtssteil, in englisch: skewed to the left 
% (psigamma(0.5, deriv = 2) - psigamma(1, deriv = 2))/(psigamma(1, deriv = 1) + psigamma(0.5, deriv = 1))^(3/2)

@@

Ireland can be considered a rather special case. It is the only economy where the effects of the ongoing financial crisis of 2008 are visible. We find a structural break in 2008 following the strong contraction triggerd by the bursting of the Irish housing bubble and a beginning deflation.

@@

Another interesting group consists of 3 economically rather different countries. In all of them we find no evidence of a structural change. Whereas the Netherlands and Finland are both on the lower side of the inflation margin, Greece exhibits some very strange behaviour. It has by far the highest variance of all series and seems to be completely different from all the others.

@@

The last group with clear patterns can be seen with the two countries that decided not to be part of any European Monetary System\footnote{If we ignore the short participating time of the United Kingdom in the EMS}, the United Kingdom and Sweden. Both have a clear break in the early 90ies, and both of these breaks can be traced back to crisis, the currency crisis of the United Kingdom cumulating into the "Black Wednesday" in 1992 and the run on the Swedish currency in the same year.

@@

No clear interpretation is available for: Austria, Denmark, France, Germany and Portugal.

@@

\subsection{An example--Slovenia}
\normalsize

Let us go through all the steps of our analysis by means of an example, for this we have chosen the case of Slovenia. The economic reason why we suppose to find changing dynamics are clear. In the first phase of the consolidation after the break--up of Yugoslavia and after the short war of 1991, Slovenia experienced very high inflation rates, 
with yearly growth rates as high as 90\% in August 1990. But gradually Slovenia was able to reduce the inflation rate towards much lower values. This first phase is not captured in our data since it starts as late as January 1996. For the period up to today we would expect one or two breaks from an economic point of view. The efforts to reach the Maastricht criteria as soon as possible are likely to lead to a further lowering of the mean and possibly also of the variance of the inflation rate. Another break point might possibly be found towards the introduction of the EURO coins themselves in 2007. \\
Actually when we look at the data, this is precisely what we get:

<<fig=TRUE, height=4, width=8>>=
x <- results$Slovenia
plot(x$data, xlab = "Time", ylab = "100 * diff(log(HICP))", main = "")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
@

<<fig=TRUE, height=4, width=8>>=
plot(x$efp, functional = supLM(0.1), main = "supLM test", xaxt = "n", xlim = range(time(x$data)))
Axis(time(x$data), side = 1)
lines(x$breakpoints, breaks = x$nbreaks)
@

The supLM test graph  suggests one break around 2003 and possibly another towards the end of the sample. This second break point is not estimated, since we restricted a regime towards a length of at least 24 month. Thus we find one break in August 2003 with a much lower mean value (only a third now of its previous value) and interestingly enough, a higher volatility.\\
To be as precise as possible with the detection of break points, we also include a graph showing the values of the LWZ (Liu, Wu and Zidek) criteria as developed in \citet{liu}. Whereas the supLM test suggest at least one break, the LWZ criterion favors no break. But the decline in the negative log--likelihood from 0 breaks to 1 is the strongest, we concluded that one break point actually would be the best alternative.

@@

Slovenia did achieve this by two very important measures. First it liberalized the financial markets in 2003 and -- most important -- it restricted the year--on--year growth of the monetary base M1 severely starting in 2003. Combined, these actions were enough to ensure  Slovenian participation in the EURO--zone.

@@

\clearpage
\section{Discussion}

% actually too detailed for so small a picture

\begin{center}
<<greyplot, fig=TRUE, height=8, width=8>>=
 ## compute GL mean (if no NA)
  glmean <- function(x) {
    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
  }
  ## returns
  ## why not include Finland?
  ## hicp <- diff(log(HICP[, -6]))
  hicp <- diff(log(HICP[,]))

  ## rolling GL means
  hicp_glmean <- rollapply(hicp, 12, glmean)

  ## semi-transparent gray and red
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)

  ## plot returns (gray) and rolling means (red)
  plot(hicp, plot.type = "single", lwd = 1.5, col = gray_red[1], xlab="Time", ylab="Series")
  for(i in 1:ncol(hicp)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\end{center}
